---
title: "DATA 304 Homework 8"
author: "Lily McAboy"
date: "2/24/2025"
format:
  html:
    embed-resources: true
    code-tools: true
---
```{r}
#| include: false
library(vegawidget)
library(dplyr)
library(reticulate)
library(altair)
library(jsonlite)
library(htmltools)
library(tidyverse)
library(tidyr)
library(vegabrite)
pd <- import("pandas")
```

```{python}
import altair as alt
import pandas as pd
import ssl
import certifi
import urllib.request
```

# Exercise 1

### Import Data

```{r}
supply_demand <- fromJSON("https://calvin-data304.netlify.app/data/swd-lets-practice-ex-2-3.json")
```

```{python}
supply_demand = pd.read_json("https://calvin-data304.netlify.app/data/swd-lets-practice-ex-2-3.json")
```

```{r}
# Wrangle the dataset to find count
long_supply_demand <- supply_demand |> 
  pivot_longer(
    cols = c(capacity, demand),
    names_to = "type",
    values_to = "value"
  )
```

```{python}
long_supply_demand = supply_demand.melt(id_vars=["date"], var_name="type", value_name="Value")
```

## Graph 1: Dodged Bar Chart

I realized that there is an issue with Python Altair. When I make date type "Temporal", it will not xOffset. If I make it ordinal, it will xOffset. 

[Python Altair Grouped Bar Graph Example](https://altair-viz.github.io/gallery/grouped_bar_chart2.html)

```{python}
alt.Chart(long_supply_demand).mark_bar().encode(
  x="date:O",
  y="Value:Q",
  xOffset="type:N",
  color="type:N"
)
```


## Graph 2: Line Chart
```{r}
s_line <- vl_chart() |>
  vl_mark_line(color = 'orange', size = 3) |>
  vl_encode_y(field = 'capacity:Q')

d_line <- vl_chart() |>
  vl_mark_line(color = 'blue', size = 3) |>
  vl_encode_y(field = 'demand:Q')

vl_layer(s_line, d_line) |>
  vl_encode_x('date:T', title = 'Date') |>
  vl_encode_y(type = 'quantitative') |>
  vl_axis_y(title = '') |>
  vl_add_properties(width = 400, height = 300) |>
  vl_add_data(supply_demand)
```


## Graph 3: Stacked Bar Chart with Different Sized Bars

```{r}
vl_chart(width = 300, height = 300) |>
  vl_mark_bar() |>
  vl_encode_x("date:T") |>
  vl_encode_y("value:Q") |>
  vl_encode_color("type") |>
  vl_encode_size(
    "type:N", 
    scale = list(
      domain = list("capacity", "demand"),
      range = list(5, 10) 
    )
  ) |>
  vl_add_data(long_supply_demand)
```

## Graph 4: Stacked Bar Chart

```{r}
vl_chart(width = 300, height = 300) |>
  vl_mark_bar() |>
  vl_encode_x("date:T") |>
  vl_encode_y("value:Q") |>
  vl_encode_color("type") |>
  vl_add_data(long_supply_demand)
```


## Graph 5: Connected Point Chart

```{python}

s_circle = alt.Chart(supply_demand).mark_circle(color='orange', size=100).encode(
    x='date:T',
    y='capacity:Q'
)

d_circle = alt.Chart(supply_demand).mark_circle(color='blue', size=100).encode(
    x='date:T',
    y='demand:Q'
)

lines = alt.Chart(supply_demand).mark_rule().encode(
    x='date:T',
    y='capacity:Q',
    y2='demand:Q'
)
chart = s_circle + d_circle + lines
chart.properties(width=400, height=300)

```

## Graph 6: Layered Point and Line Chart
```{r}
d_point <- vl_chart() |>
  vl_mark_point(color = 'blue', size = 20) |>
  vl_encode_y(field = 'demand:Q')

vl_layer(d_point, d_line) |>
  vl_encode_x('date:T', title = 'Date') |>
  vl_encode_y(type = 'quantitative') |>
  vl_axis_y(title = '') |>
  vl_config_title('2019 Supply and Demand') |>
  vl_add_properties(width = 400, height = 300) |>
  vl_add_data(supply_demand)
```


# Exercise 2

### a. There was some question as to just what perc meant in this data set. Create a graphic that shows the sum of all the values of perc in each year separated by men and women. Explain what your graphic reveals about what perc means.

```{python}
jobs = pd.read_json("https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json")
```

Python Altair gives this error if the data set is too large: ! altair.utils.data.MaxRowsError: The number of rows in your dataset is greater than the maximum allowed (5000).

So for the sake of the exercise, I will filter out rows where count = 0 and years greater than 1870.  
```{python}
filtered_jobs = jobs[(jobs['count'] != 0) & (jobs['year'] > 1870)]
```

```{python}
alt.Chart(filtered_jobs, title="Sum of Gender Differences For All Jobs").mark_bar().encode(
    alt.X('year:O').title('Year'),
    alt.Y('sum(perc):Q').title('Percentage'),
    alt.Color('sex:N'),
    alt.Column('sex:N').title('')
)
```

The graphic shows the aggregated sum for each percentage for each year per sex. For example, all of the jobs for 1880 for men is around 85%. This means that the work demographic for men was ~85% in 1880, and women made up ~15% of the workforce. For each year, the addition of the female and male percentages should equal 100%. 

### b. Now modify the graphic from class so that there are facets for 4 different years (your choice which years), scatterplots that show the percent of men and of women who worked in various jobs during those years, and format=".0%" is used (in the appropriate place) to display perc as a percentages (multiplied by 100 with a % sign appended).

### Facets

```{r}
jobs <- fromJSON("https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json")
```

### Here is a gross amount of data wrangling that I had to do to put it into the data formatting I wanted
```{r}
filtered_year_jobs_wide <- jobs |>
  select(job, sex, year, perc) |>
  pivot_wider(names_from = year, values_from = perc, names_prefix = "perc_") |>
  pivot_longer(cols = starts_with("perc_"), 
               names_to = "compare_year", 
               values_to = "perc_other",
               names_prefix = "perc_") |>
  filter(compare_year != "2000") 

perc_2000_data <- jobs |>
  filter(year == 2000) |>
  select(job, sex, perc) |>
  rename(perc_2000 = perc)

jobs_comparison <- jobs |>
  filter(year != 2000) |>
  left_join(perc_2000_data, by = c("job", "sex")) |> 
  rename(perc_other = perc)

jobs_comparison <- filter(jobs_comparison, year %in% c(1850, 1900, 1950, 1970, 2000))

jobs_comparison <- jobs_comparison |> 
  mutate(perc_other = perc_other * 100, 
         perc_2000 = perc_2000 * 100)
```

```{r}
vl_chart() |>
  vl_mark_circle() |>
  vl_encode_x("perc_other:Q", title = "Percentage For Year") |> 
  vl_scale_x(type = "symlog", constant = 0.00001) |>  
  vl_encode_y("perc_2000:Q", title = "2000 (%)") |> 
  vl_scale_y(type = "symlog", constant = 0.00001) |>  
  vl_facet_column("year:N", title = "") |>
  vl_encode_tooltip_array(list("job", "sex", "compare_year", "perc_other", "perc_2000")) |>  
  vl_add_properties(
    title = list( 
      text = "Comparison of Occupation Percentages: 2000 vs Other Years",
      subtitle = "Each dot represents an occupation. Hover to see details."
    )) |>
  vl_add_data(jobs_comparison)

```

### c. Redo part b but show the percentage of men who worked in various jobs and the percentage of women who worked in various jobs. (Part a should have convinced you that this is NOT what perc is in the original data set.)

I am a bit confused by this question so I will make a couple of graphs to show what I thought it meant. 

```{r}
vl_chart() |>
  vl_mark_circle() |>
  vl_encode_x("perc_other:Q", title = "Percentage For Year") |> 
  vl_scale_x(type = "symlog", constant = 0.00001) |>  
  vl_encode_y("perc_2000:Q", title = "2000 (%)") |> 
  vl_scale_y(type = "symlog", constant = 0.00001) |>  
  vl_facet_column("year:N", title = "") |>  
  vl_encode_color("sex:N", title = "Sex") |>
  vl_scale_color(
    domain = list("men", "women"),
    range = list("lightblue", "pink")
  ) |>
  vl_encode_tooltip_array(list("job", "sex", "compare_year", "perc_other", "perc_2000")) |>  
  vl_add_properties(
    title = list( 
      text = "Comparison of Occupation Percentages: 2000 vs Other Years",
      subtitle = "Each dot represents an occupation. Hover to see details."
    )) |>
  vl_add_data(jobs_comparison)

```

```{r}
jobs_sex <- jobs |>
  pivot_wider(names_from = sex, values_from = perc)
```

```{r}
vl_chart() |>
  vl_mark_point() |>
  vl_pivot("sex", groupby = list("job", "year"), value = "perc") |>
  vl_encode_x("women:Q") |>
  vl_scale_x(type = "symlog", constant = 0.00001) |>
  vl_encode_y("men:Q") |>
  vl_encode_color("year:O") |>
  vl_scale_y(type = "symlog", constant = 0.00001) |> 
  vl_add_properties(
    title =  list( 
      text = "Percent of people working various jobs"
      )) |>
  vl_add_data(jobs) 
```

I think that the blue scale for year is kind of confusing - but I do see a trend that a lot of light colored (earlier years) dots go up along the y axis, which shows that earlier centuries had less women in the workforce than men. 

# Exercise 3

### Read in Data

```{python}
gas_prices = pd.read_csv("https://calvin-data304.netlify.app/data/pump_price_for_gasoline_us_per_liter.csv")
country_codes = pd.read_csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv")

country_codes.rename(columns={"name": "country"}, inplace=True)
```

### Left Outer Join 
```{python}
price_and_code = country_codes.merge(gas_prices, on="country", how = "left")
```

I first tried left joining country codes onto gas prices, but since gas prices only had 182 rows and country codes has 249 rows, I switched them so the new price_and_code merged set doesn't "lose" any of the countries. Though, this also means that there were a good chunk of countries that either didn't match up so may have nulls. If I inspect my dataset, I can see that West Sahara is an example where we joined it to gas prices but there is no data for gas prices. 

### Graphics
```{python}
gas_prices_long = gas_prices.melt(id_vars=["country"], var_name="Year", value_name="Price")
price_and_code_long = country_codes.merge(gas_prices_long, on = "country", how = "left")
```

### Graph 1: Change in Gas Prices By Region By Year
```{python}
alt.Chart(price_and_code_long, title = "Gas Prices Per Region").mark_circle(size=100).encode(
  x = alt.X("Year:T", title = "Year"),
  y = alt.Y("average(Price):Q", title = "Price in US Dollars"),
  color = alt.Color("region:N")
).properties(
    width=600,
    height=400
)
```

### Graph 2: Heat Map By Year By Sub-Region

```{python}
alt.Chart(price_and_code_long, title = "Gas Prices Per Region").mark_rect().encode(
  y=alt.Y('sub-region:N', title= '', sort='-color'),
  x=alt.X('Year:O', title = 'Year', axis=alt.Axis(labelAngle=0)),
  color=alt.Color('Price:Q',title = 'Price', scale=alt.Scale(scheme='inferno'))
).configure_axis(
    grid=True,
    tickBand='extent'
).properties(
    width=500,
    height=500
)
```

### Graph 3: 

```{python}
alt.Chart(price_and_code, title = "Sub-Region Gas Prices 2000 vs. 2016").mark_circle(opacity=0.7, size = 100).encode(
    x=alt.X('2016:Q', title='Price in 2016 (USD)'),
    y=alt.Y('2000:Q', title='Price in 2000 (USD)'),
    color='sub-region:N',
    tooltip=['country:N', 'region:N', 'alpha-3:N', '2000:Q', '2016:Q']
).properties(
    width=500,
    height=600
)
```

